---
output: html_document
---
# Data Preparation and Association Rule Mining

#### Reading Employee Attrition Data

```{r data, message=F, warning=F}
library(tidyverse)
emp_atr <- read_csv("employee_attrition.csv", col_types = cols())
attr(emp_atr, 'spec') <- NULL
dim(emp_atr)
```

```{r structure}
str(emp_atr)
```

#### Preview of Data

```{r preview, message = F, warning = F}
library(knitr)
knitr::kable(head(emp_atr))
```

Removing those attributes that are not useful for the analysis

```{r rmatr}
emp_atr <- select(emp_atr, -c("EmployeeCount", "EmployeeNumber", "Over18", "StandardHours"))
```

#### Changing Attribute Type

Converting character type to factor

```{r factors}
char_var <- sapply(emp_atr, is.character)
emp_atr[,char_var] <- lapply(emp_atr[,char_var], as.factor)
```

### Data Quality Issues

#### 1. Data outliers

Looking for outliers by plotting a boxplot

```{r outlier1}
boxplot(emp_atr$TotalWorkingYears)
```

```{r outlier2}
boxplot(emp_atr$YearsWithCurrManager)
```

Omitting the extreme outliers in total working years and years with current manager as the values are much higher than corresponding employee age

```{r rmoutlier}
outlier_obs <- c(which(emp_atr$TotalWorkingYears == 114), which(emp_atr$YearsWithCurrManager == 219))
emp_atr <- emp_atr[-outlier_obs, ]
```

#### 2. Missing values

Checking for number missing values

```{r nas}
sum(is.na(emp_atr))
```

Looking for attributes with missing valies

```{r naatr}
which(apply(emp_atr, 2, function(x) sum(is.na(x))) > 0)
```

Replacing missing value with mean of the attribute

```{r nareplace}
emp_atr$DistanceFromHome[is.na(emp_atr$DistanceFromHome)] <- mean(emp_atr$DistanceFromHome, na.rm = T)
emp_atr$JobLevel[is.na(emp_atr$JobLevel)] <- mean(emp_atr$JobLevel, na.rm = T)
emp_atr$PercentSalaryHike[is.na(emp_atr$PercentSalaryHike)] <- mean(emp_atr$PercentSalaryHike, na.rm = T)
emp_atr$PerformanceRating[is.na(emp_atr$PerformanceRating)] <- mean(emp_atr$PerformanceRating, na.rm = T)
emp_atr$RelationshipSatisfaction[is.na(emp_atr$RelationshipSatisfaction)] <- mean(emp_atr$RelationshipSatisfaction, na.rm = T)
emp_atr$TotalWorkingYears[is.na(emp_atr$TotalWorkingYears)] <- mean(emp_atr$TotalWorkingYears, na.rm = T)
emp_atr$YearsSinceLastPromotion[is.na(emp_atr$YearsSinceLastPromotion)] <- mean(emp_atr$YearsSinceLastPromotion, na.rm = T)
```

Since both Gender and OverTime are of type factor, it is better to omit the observations with missing values for further analysis

```{r naomit}
emp_atr <- emp_atr[-c(which(is.na(emp_atr$Gender)), which(is.na(emp_atr$OverTime))),]
ea <- emp_atr
```

#### 3. Duplicate records

Counting number of duplicate records  in the data

```{r duplicate}
nrow(emp_atr[duplicated(emp_atr), ])
```

### Exploratory Data Analysis

Comparing employee rate and income based on different education field

```{r eda1}
library(dplyr)
kable(emp_atr %>% group_by(EducationField) %>%
  summarise(TotalEmployees = n(), DailyRate = round(mean(DailyRate, na.rm = T)), 
            HourlyRate = round(mean(HourlyRate, na.rm = T)),
            MonthlyIncome = round(mean(MonthlyIncome, na.rm = T)),
            MonthlyRate = round(mean(MonthlyRate, na.rm = T)),
            PercentSalaryHike = round(mean(PercentSalaryHike, na.rm = T))) %>%
  arrange(desc(TotalEmployees)))
```

Comparing employee rate and income in each department based on gender 

```{r eda2}
kable(emp_atr %>% group_by(Department, Gender) %>%
  summarise(TotalEmployees = n(), DailyRate = round(mean(DailyRate, na.rm = T)), 
            HourlyRate = round(mean(HourlyRate, na.rm = T)),
            MonthlyIncome = round(mean(MonthlyIncome, na.rm =T)),
            PercentSalaryHike = round(mean(PercentSalaryHike, na.rm = T))))
```

Comparing employees with different job role based on monthly income and percentage hike in salary

```{r eda3}
kable(emp_atr %>% group_by(JobRole) %>%
  summarise(TotalEmployees = n(), MonthlyIncome = round(mean(MonthlyIncome, na.rm =T)),
            PercentSalaryHike = round(mean(PercentSalaryHike, na.rm = T))) %>%
  arrange(desc(TotalEmployees)))
```

Determining employee attrition based on gender

```{r eda4}
kable(emp_atr %>% group_by(Gender, Attrition) %>%
  summarise(TotalEmployees = n()))
```

Determining employee attrition based on job role

```{r eda5}
kable(emp_atr %>% group_by(JobRole, Attrition) %>%
  summarise(TotalEmployees = n()))
```

Determining employee attrition based on working overtime

```{r eda6}
kable(emp_atr %>% group_by(OverTime, Attrition) %>%
  summarise(TotalEmployees = n()))
```

### Dimension Reduction

Determining correlation between different attributes

```{r corr, warning=F, message=F}
library(corrplot)
cor_matrix <- cor(emp_atr[complete.cases(emp_atr), sapply(emp_atr, is.numeric)], method = "pearson")
corrplot(cor_matrix, type = "upper")
```

Removing attributes that have very low correlation

```{r dimreduce}
emp_atr <- select(emp_atr, -c("Education", "DailyRate", "DistanceFromHome", "EnvironmentSatisfaction", "HourlyRate", 
                              "JobInvolvement", "JobSatisfaction", "MonthlyRate", "NumCompaniesWorked",
                              "PercentSalaryHike", "PerformanceRating", "RelationshipSatisfaction", "StockOptionLevel", 
                              "TrainingTimesLastYear", "WorkLifeBalance"))
```
### Discretization

Converting numerical attributes to categorical as Association Rule Mining works only with categorical variables

```{r discretize, message = F, warning = F}
library(arules)
library(arulesViz)
emp_atr$Age <- arules::discretize(emp_atr$Age, method = "fixed", 
                            breaks = c(18, 30, 45, 60), labels = c("Below 30", "Between 30 and 45", "Above 45"))
emp_atr$JobLevel <- arules::discretize(emp_atr$JobLevel, method = "interval",
                            breaks = 5, labels = c("1", "2", "3", "4", "5"))
emp_atr$MonthlyIncome <- arules::discretize(emp_atr$MonthlyIncome, method = "fixed",
                            breaks = c(0,5000, 10000, 15000, 20000), labels = c("Less than 5000", "Between 5000 and 10000",
                            "Between 10000 and 15000", "Above 15000"))
emp_atr$TotalWorkingYears <- arules::discretize(emp_atr$TotalWorkingYears, method = "interval",
                            breaks = 3, labels = c("Less", "Medium", "High"))
emp_atr$YearsAtCompany <- arules::discretize(emp_atr$YearsAtCompany, method = "interval",
                            breaks = 3, labels = c("Less", "Medium", "High"))
emp_atr$YearsInCurrentRole <- arules::discretize(emp_atr$YearsInCurrentRole, method = "interval",
                            breaks = 3, labels = c("Less", "Medium", "High"))
```

### Visualization

#### Department vs Attritiom

```{r bar, message = F, warning = F}
library(ggplot2)
ggplot(ea, aes(x = Department, fill = Attrition)) +
  geom_bar()
```

Most of the employees who leave the company belong to Research and Development department

#### MonthlyIncome vs Attrition

```{r box}
ggplot(ea, aes(x = Attrition, y = MonthlyIncome)) +
  geom_boxplot()
```

Employees with less monthly income choose to leave the company

#### MonthlyIncome vs YearsInCurrentRole

```{r scatter}
ggplot(ea,aes(x = YearsInCurrentRole, y = MonthlyIncome)) +
  geom_point(aes(color = Gender, size = YearsAtCompany)) +
  geom_smooth(method = "lm")
```

Male employees tend to earn more than female employees

#### JobLevel density and Attrition

```{r density}
ggplot(ea, aes(JobLevel)) +
  geom_density(aes(fill = Attrition)) +
    labs(subtitle = "Job level grouped by Attrition", x="JobLevel",fill="Attrition")
```

Employees at lower job levels leave the company more often than the employees at higher level

### Association Rule Mining

#### Baseline model

Generating transcational dataset from record dataset

```{r transaction, message = F, warning = F}
fac_var <- sapply(emp_atr, is.factor)
employee <- as(emp_atr[, fac_var], "transactions")
inspect(head(employee, 2))
```

Running Apriori on transactional dataset with default values

```{r base, results = "hide"}
rules_transaction <- apriori(employee)
```

```{r i1}
inspect(head(rules_transaction, 5))
```

Fine tuning the baseline model by assigning support and confidence values

```{r tuning1, warning = F, results = "hide"}
rules_transaction <- apriori(employee, parameter = list(support = 0.01, confidence = 0.3, minlen = 3))
```

```{r i2}
inspect(head(rules_transaction, 5))
```

- Very low value of support will simply include the rules that occoures by chance

- Low confidence imples the inference made by the rule is not very reliable

- Having low value of both support and confidence will lead to overfitting and produce a lot more rules than necessary

Increasing the value of support and confidence and running the model

```{r tuning2, results = "hide"}
rules_transaction <- apriori(employee, parameter = list(support = 0.5, confidence = 0.95, minlen = 3))
```

```{r i3}
inspect(head(rules_transaction, 5))
```

- We might miss out on some interesting rules if the value of support if very high

- Having very high confidence will lead to underfitting ang generate fever rules than expected

- Fine tuning the model with appropriate value of support and confidence is very essential in order to capture significant rules whose inference can be reliable

Setting support to 0.1, confidence to 0.85 and inspecting the rules

```{r tuning3,results = "hide"}
rules_transaction <- apriori(employee, parameter = list(support = 0.1, confidence = 0.85, minlen = 3))
```

```{r i4, warning = F}
inspect(head(rules_transaction, 5))
```

#### Predicting Attrition

##### 1. Attrition = Yes

``` {r leave}
rules1 <- apriori(data = employee, parameter = list(supp = 0.01, conf = 0.5, minlen = 3),
                 appearance = list(default = "lhs", rhs = "Attrition=Yes"),
                 control = list(verbose = F))
```

Removing redundant rules

```{r leaveredundant}
subset_rules <- which(colSums(is.subset(rules1, rules1)) > 1)
rules1 <- sort(rules1[-subset_rules], by = "lift", descreasing = T)
```

Top 5 rules for employees leaving the company

```{r inspectleave}
inspect(head(rules1, 5))
```

- From the top five rules it is evident that employees who often work overtime in sales department tend to leave the company
 
- These rules are significant as they have high lift and confidence
 
- Interesting rules are the ones in which single employees choose to leave the company

Plot for rules with Attrition=Yes

```{r leavescatter}
plot(rules1, measure = c("support", "lift"), shading = "confidence", jitter = 0)
```

##### 2. Attrition = No

``` {r stay}
rules2 <- apriori(data = employee, parameter = list(supp = 0.1, conf = 0.5, minlen = 3),
                 appearance = list(default = "lhs", rhs = "Attrition=No"),
                 control = list(verbose = F))
```

Removing redundant rules

```{r stayredundant}
subset_rules <- which(colSums(is.subset(rules2, rules2)) > 1)
rules2 <- sort(rules2[-subset_rules], by = "lift", descreasing = T)
```

Top 5 rules for employees staying with the company

```{r stayinspect}
inspect(head(rules2, 5))
```

- From the top 5 rules it is evident that employees who are in current role for quite some years tend to stay with the company

- The interesting rules are the ones in which divorced employees choose to stay win the company

Plot for rules with Attrition=No

```{r stayscatter}
plot(rules2, measure = c("support", "lift"), shading = "confidence", jitter = 0)
```

### Shiny App

Link: [Association Rule Mining](https://nishithavenkatesh.shinyapps.io/associationrule/)